Municipalities dataset cleaning

### 2023-01-11 data_cleaning/Data_cleaning_and_covariates_Jan_09.qmd
Added data for the health region approach

We have obtained and cleaned the data of the municipalities from the province of Ontario website, and cleaned the dataset to obtain the city names and geographical locations. Further details can be found in `municipalities.qmd` in the `data_cleaning` directory. Next, we will join and match the geographical location from the municipalities dataset to the clean dataset we have obtained in this document so far after removing the entries that do not have a corresponding geographical location.

@@

There are 756 observations that did not get a geographical region, and they correspond to 177 unique cities. These will be exported to a csv file in order to manually write their geographical areas following the Association of Municipalities of Ontario divisions, and using Wikipedia to check the status of each municipality.

@@

geographic_areas$Geographic.area<-as.factor(geographic_areas$Geographic.area)

geographic_areas$Geographic.area.title <-as.factor(geographic_areas$Geographic.area.title)

@@

clean_data <-left_join(clean_data,geographic_areas,by="Geographic.area")

@@

**This section needs to be modified as the correction has been done by geographica area on Jan 8, 2023**

@@

>>>>

We have obtained and cleaned the data of the municipalities from the province of Ontario [website](https://www.ontario.ca/page/list-ontario-municipalities#section-3), and cleaned the dataset to obtain the city names and geographical locations. Further details can be found in `municipalities.qmd` in the `data_cleaning` directory. We will join and match the geographical location from the municipalities dataset to the clean dataset we have obtained in this document so far after removing the entries that do not have a corresponding geographical location.

@@

There are 2744 observations that did not get a geographical region, and they correspond to 187 unique cities. These will be exported to a csv file in order to manually write their geographical areas following the Association of Municipalities of Ontario divisions, and using Wikipedia to check the status of each municipality.

@@

geographic_areas$Geographic_area<-as.factor(geographic_areas$Geographic_area)

geographic_areas$Geographic_area_title <-as.factor(geographic_areas$Geographic_area_title)

@@

clean_data <-left_join(clean_data,geographic_areas,by=c("Geographic.area"="Geographic_area")) %>%
   subset(select=-c(Full_title.y))

@@

**This section needs to be modified as the correction has been done by geographical area on Jan 8, 2023**

@@

# Health Regions approach

Following on a suggestion by BN on Jan 09, the idea is to use broader regions for geographical analysis. The Health Regions of Ontario will be used. However, these Health Regions do not match the census divisions. I have not been able to find a dataset from Health Ontario that lists each municipality and its corresponding Health Region. I will use a multi-stage approach to incorporate the information into the dataset:

- Get dataset from Paul Allen regarding long-term care homes in Ontario (https://paulallen.ca/consolidated-dataset-of-ltc-homes-in-ontario/). This dataset has communities and the Local Health Integration Network (LHINs) where long-term care homes were located. I will keep the community and LHIN information.

- Add column for Health Region based on LHIN: The Health Regions substituted LHINs, a list of their correspondence can be found here: https://www.ontariohealth.ca/about-us/our-people. I will add this information.

- Merge the dataset and check if there are missing municipalities. Manually add missing municipalities.

## Steps

The dataset from Paul Allen was downloaded and saved as `Consolidated_LTC_dataset.csv` in the `data` directory. One thing to note is that there is a missing observation (coded as "Not provided") for one of the entries (city of Napanee, located in the Lennox and Addington County). Under the LHIN divisions, Napanee was in the South East LHIN. Also, there is one entry that says "244 Main Street East" as the community entry but it should be "Stayner" (the address provided belongs to Stayner.) The information is fixed before adding the Health Region.

```{r,get-health-data}

#load data
ltc_data <- read.csv(here("data","Consolidated_LTC_dataset.csv"))

## checking missing observations

sum(ltc_data$X_LHIN=="Not provided")

ltc_data <- ltc_data %>%
  mutate(X_LHIN=replace(X_LHIN,X_LHIN=="Not provided","South East"))

ltc_data <- ltc_data %>%
  mutate(COMMUNITY=replace(COMMUNITY,COMMUNITY==" 244 Main Street East","Stayner"))

#keep relevant columns and add Health region information
ltc_data <-ltc_data %>%
 select(c("COMMUNITY","X_LHIN"))%>%
  rename(city=COMMUNITY,
         LHIN=X_LHIN)%>%
  mutate(Health_Region=
    case_when(LHIN=="Central"~ "Central",
              LHIN=="Central West"~"Central",
              LHIN=="Mississauga Halton"~"Central",
              LHIN=="North Simcoe Muskoka"~"Central",
              LHIN=="Central East"~"East",
              LHIN=="South East"~"East",
              LHIN=="Champlain"~"East",
              LHIN=="North East"~"North East",
              LHIN=="North West"~"North West",
              LHIN=="Toronto Central"~"Toronto",
              LHIN=="South West"~"West",
              LHIN=="Hamilton Niagara Haldimand Brant (Hnhb)"~"West",
              LHIN=="Waterloo Wellington"~"West",
              LHIN=="Erie St. Clair"~"West"
    )
  )%>%
  distinct(city,.keep_all = TRUE)

Now, combine datasets and write csv file for those cities that were not assigned a Health Region.

```{r,combine}

clean_data<-left_join(clean_data,ltc_data,by="city")

 missing_health_regions<-clean_data %>% filter(is.na(LHIN)) %>%
   distinct(city,.keep_all = TRUE)%>%
   select(city,Geographic.area,Geographic_area_title,LHIN,Health_Region)

#write.csv(missing_health_regions,here("data","missing_health_regions.csv"),row.names = FALSE)

To complete the missing LHIN information, the correspoding LHINs were obtained from the LHIN websites:

https://www.southeasthealthline.ca/aboutourregion.aspx?region=RuralFrontenacLennoxAddington

https://www.nsmhealthline.ca/aboutourregion.aspx

https://www.champlainhealthline.ca/aboutourregion.aspx?region=StormontDundasGlen

https://www.wwhealthline.ca/aboutourregion.aspx?region=KitchenerWaterlooWellesleyWilmotWoolwich

Some cities did not belong entirely to a LHIN. For example, Etobicoke was divided between the Central, Central East, and Toronto Central LHINs. We chose in these cases the LHIN that covered the larger geographical region of each city (Toronto Central LHIN in the case of Etobicoke). We now assign LHINs to the dataset and Health Regions. The dataset with the information of the missing LHINs is called `missing_health_regions_updated.csv`

```{r,merge-health-regions}

mhru<-read.csv(here("data","missing_health_regions_updated.csv"))

### 2023-01-11 data_cleaning/municipalities.qmd
Added data for the health region approach

The file obtained is a CSV but it needed to be cleaned to remove the innecesary information found in the name of each city so it can be combined with the survey dataset.

```{r,loading}
#| echo: false
#| warning: false

@@

municipalities_clean <- municipalities %>%
#get only the city part (see https://stackoverflow.com/questions/66212344/extract-word-in-quotes-from-string)

  mutate(city=str_extract(Municipality,'(?<=(?<!\\()")[^"]+(?=(?!\\))")'))%>%
#keep only the name of the city
mutate(city=word(city,sep=",")) %>%
  #remove the first column as it is no longer needed
  subset(select=-c(Municipality))

@@

One additional step is to remove from the dataset duplicate entries. In the dataset there are some instances that appear as upper tier division (counties, regions) that have the same name of a municipality, thus causing duplicates. The ones to remove are identified below.

@@

municipalities_clean[duplicated(municipalities_clean$city),]

@@

R detected the "duplicate" entries because the names of the cities we are interested in sometimes appear after the entry we want to remove. After examining the original municipalities dataset (`municipalities_list.csv`), and comparing it with the list provided above, the duplicate entries in the dataset that need to be removed correspond to:

-   County of Essex
-   Township of Hamilton (it is an aggregation according to https://en.wikipedia.org/wiki/Hamilton_Township,\_Ontario)
-   County of Perth
-   County of Peterborough
-   County of Renfrew
-   Regional Municipality of Waterloo

These entries need to be removed as we are interested in municipalities within larger geographical areas, and these are aggregations that do not fit that criteria. The final step is to remove the duplicates. Because the word that identifies them as aggregations (i.e., "County", "Township") is lost in the process of getting the names of the cities to match them with the survey dataset, we will remove them by rownumber (after comparing with the `municipalities_list.csv`) dataset. The rownumbers are:

-   County of Essex: 119
-   Township of Hamilton: 155
-   County of Perth: 313
-   County of Peterborough: 317
-   County of Renfrew: 338
-   Regional Municipality of Waterloo: 420

```{r,save-file}

#| echo: false

municipalities_clean <- municipalities_clean[-c(119,155,313,317,338,420),]

#save the file in csv
write.csv(municipalities_clean,here("data","municipalities_clean.csv"),row.names = FALSE)
```

>>>>

The file obtained is a CSV but it needed to be cleaned to remove the unnecessary information found in the name of each city so it can be combined with the survey dataset.First, we only keep the name of the city and the municipal region they belong to, and later remove duplicate entries.

```{r,loading-first-clean}
#| echo: false
#| warning: false

@@

municipalities_clean <- municipalities %>%
#get only the city/county part (see https://stackoverflow.com/questions/66212344/extract-word-in-quotes-from-string)
  mutate(Full_title=str_extract(Municipality,'(?<=(?<!\\()")[^"]+(?=(?!\\))")'))

@@

The second step of cleaning this dataset is the fact that it contains the names of the regions (e.g., the Regional Municipality of Peel appears listed along the geographical area of Peel, when they are the same thing). These entries need to be removed as in some cases some towns have the same name as a geographical region (e.g., Durham in Grey County, having the same name as the Regional Municipality of Durham). To do this, we will use the geographical regions listed on the [website](https://www.amo.on.ca/about-us/municipal-101/ontario-municipalities) of the Association of Municipalities of Ontario (AMO). As a clarification, because the "Districts" listed in this website do not serve any municipal government purpose, except on the case of the District Municipality of Muskoka only this last one needs to be removed from the municipality list. Lastly, on the AMO website some municipalities have their most recent status (e.g., Brant, which appears as city, but still appears listed as "County") in the website of the province of Ontario. We will load the AMO dataset (called `geographic_areas.csv`) and use to remove those entries from `municipalities_clean` that match the entries. Later, we will manually remove entries that do not have the same name but correspond to the same region.

```{r,clean-second-step}

# load AMO dataset
geographic_areas <- read.csv(here("data","geographic_areas.csv"))

#remove entries with the same name in the Ontario dataset

municipalities_clean <- anti_join(municipalities_clean,geographic_areas,by="Full_title")

# manually remove entries

exclude <-tibble(Full_title=c("Brant, County of", 
                                       "Haldimand County", 
                                       "Norfolk County",
                                       "Prince Edward, County of",
                                       "Chatham-Kent, Municipality of"))

 #remove entries

municipalities_clean <-anti_join(municipalities_clean,exclude, by="Full_title")%>%
  #remove the first column as it is no longer needed
  subset(select=-c(Municipality))

One additional step is to create a column that only has the city name (as the Fields Dataset only has the city name on it without the full name of "Township of", or "City of"). After that, the dataset is saved as a csv file.

@@

municipalities_clean <- municipalities_clean %>%
  mutate(city=sub(",.*","",Full_title))

#save the file in csv
write.csv(municipalities_clean,here("data","municipalities_clean.csv"),row.names = FALSE)

@@
